Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-20 main.tex

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}
This section discusses how to efficently map the two levels of parallelism
available in iterative face alignment to the two levels of concurrency
available in a typical multicore CPU architecture.

We will show that, in constrast with previous implementations, which mapped
problem level parallelism onto core-level concurrency, it is more efficient to
completely disregard problem-level parallelism, and instead map pixel-level
parallelism onto both core-level and sse-level cpu concurrency.  In other
words, the cpu implementation we advocate performs alignment of the test image
against the gallery images of a single user at a time.  Each matrix-vector and
vector-vector operation gets broken into 8 parts that run one per core, and
into 4 single float vectors that execute on the SSE SIMD units.
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_cpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_cpu}
\caption{Proposed mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_cpu}
\end{figure}

Recall that the L2 cache can old up to 128 image vectors' worth of data.  Since
we're only aligning data for one gallery user at a time the working set of the
problem fits into L2 cache.  Due to the higher bandwidth of L2 cache relative
to L3 cache, this results in significantly higher image alignment speed than
previous cpu implementations, as will be seen and discussed further in section
\ref{sec:alignment_benchmark}.

{\bf Implementation Details} In our implementation, most of the operations
required by the alignment stage take advantage of Intel Math Kernel Library
(MKL), a commercial implementation of the standard Basic Linear Algebra
Subprograms (BLAS).  For image resampling, we make use of Intel's Integrated
Performance Primitives (IPP) library.  Both of these libraries are heavily
optimized for Intel multicore CPU's, and are able to make good use of both
core-level and sse-level concurrency.  For operations that do not map well
onto existing libraries, we achieve both sse-level and core-level
concurrency through a combination of the OpenMP API \cite{dagum2002openmp} and
the automatic vectorization facilities of Intel's C++ compiler
(ICC)\cite{dulong1999overview}.

{\bf Comparison to Previous Implementations} The implementation presented in
\cite{WagnerA2009-CVPR} and \cite{WagnerA2011-PAMI} schedules instances of the
alignment problem to the cpu cores in a 1:1 fashion.  
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_old}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_old}
\caption{Old mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_old}
\end{figure}

This implementation was written using the OpenCV library
as a wrapper for MKL, and, like our proposed implementation, achieves sse-level concurrency implicitly.  
With eight alignment problems (associated with 8 training users) being solved
concurrently, and with 20-40 images per training user, all of the required data
fits into L3 cache.  While the L3 cache has much higher bandwidth and lower latency
than off-chip memory, this implementation is slower than the proposed implementation.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
This section discusses how to efficently map the two levels of parallelism
available in iterative face alignment to the two levels of concurrency
available in the NVIDIA Fermi GPU architecture.

Our proposed GPU implementation for the iterative alignment stage consists of a
single kernel that performs all of the test image alignment problems.  Each of
instance of the alignment problem is assigned its own thread block, and the GPU
hardware schedules as many of these thread blocks to run concurrently as the 
hardware will allow.  
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_gpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}
The full resolution test image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
recomputation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

\subsection{Other Implementation Strategies}
Several GPU implementation strategies were explored during the course of our
development, and are only mentioned briefly.  Listed in order of increasing
performance based on a synthetic benchmark of the $\ell_1$ minimization step,
they included:
\begin{enumerate}
\item Solving a single problem instance at a time using Cublas\footnote{Nvidia provided BLAS implementation} (302 ms)
\item Pure CUDA C implementation solving one problem per SMP using cuda streams\footnote{Streams is a CUDA feature that allows multiple
kernels to be run concurrently} (70 ms)
\item Pure CUDA C implementation solving one problem per SMP using one large kernel call (36 ms)
\end{enumerate}

We attribute the low performance of Cublas in this benchmark to the relatively
low problem size. Cublas is intended to take advantage of the resources of the
entire gpu to solve a single large problem, and our problem size is too small
to hide the memory latency.  Effects of this will be seen again in the
benchmarks in section \ref{sec:simulation}.

One possible implementation of the iterative alignment stage would be to
perform the optimization for all instances simultaneously, but to perform each
linear algebraic operation in a separate gpu kernel call.  The primary
advantage of this implementation strategy is that every kernel would be
performing a very simple operation requiring fewer registers per thread.  The
registers in a GPU are shared between all executing threads thus, barring other
hardware limitations, the fewer registers each thread requires, the more
threads can be scheduled to run simultaneously.  The primary downside of this
approach is that it forces synchronization between all problem instances.  It will
also have increased kernel launch overhead and potentially less efficient use of
the GPU's L2 cache.

@@

Compared to a typical CPU, a GPU has much higher peak computational bandwidth
as well as data bandwith to DRAM.  However, since our application is primarily
of matrix-vector operations and image resampling operations, it has a very low
arithmetic intensity (i.e. we perform a very small number of arithmetic
operations for every number that has to be loaded), and thus performance is
typically constrained by memory (and cache) bandwidth.  At large problem sizes
the CPU is limited by its relatively low bandwidth to DRAM compared to the gpu,
while at small problem sizes the GPU is limited by the relatively high latency
of its access to DRAM.

>>>>

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}

Optimal implementation of Algorithm \ref{alg:alignment_alm} on a multicore CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multicore CPU.

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $A$ and
$A^\dagger$ during the iterative process to execute Algorithm
\ref{alg:alignment_alm}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead map the pixel-level parallelism
onto both the core-level and the sse-level concurrency provided by the cpu.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.
This mapping of the problem is illustrated in Figure \ref{fig:alignment_mapping_cpu}.

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multicore CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. We shall compare the
performance of the two parallelism methods in Section \ref{sec:experiment}.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes it's ability to sustain a very
high bandwidth to DRAM.  A higher overall bandwidth can be achieved if 
many memory transactions are performed simultaneously, than can be achieved
with a large number of smaller transactions that are just large enough to fill
the L2 cache.

Therefore, whereas on the CPU we advocate performing alignment for one training
class at a time, on the GPU we advocate performing alignment for many classes
concurrently.  \footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is chosen by the gpu hardware. In our
experiments the hardware is able to schedule 4-5 classes per SMP to run
concurrently. This mapping of the problem-level parallelism onto the SMP-level
concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

In addition, the full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

In one experiment, we have exploited several possible GPU implementation
strategies, which include solving single alignment problems one at a time using
CUBLAS (302\,ms); solving one problem per SMP using CUDA streams (70\,ms); and
solving one problem per SMP using one large kernel call (36\,ms). The inferior
performance of CUBLAS may be attributed to the relatively small problem size.
In general, the libraries provided by Nvidia are optimized for problem sizes
that are significantly larger than our individual alignment problems.  In
Section \ref{sec:experiment}, we show the performance of our CPU and GPU
implementations running on synthetic and real image data.

@@
